Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-03-26 2024pipit/section/60-Evaluation.tex
evaluation

We were able to encode equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
For example, the specification-only function that finds the next trigger is naturally recursive.
Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
For example, given a hypothetical recursive definition of @next@ (left), the array-scan encoding is (right):

\begin{minipage}{0.4\textwidth}
\begin{verbatim}
next(i) =
  if enabled(triggers, i, c)
  then i
  else if i >= LENGTH - 1
  then NO_NEXT_TRIGGER
  else next (i + 1);
\end{verbatim}
\end{minipage}
\begin{minipage}{0.6\textwidth}
\begin{verbatim}
next_array[i] =
  if enabled(triggers, LENGTH - 1 - i, c)
  then (LENGTH - 1 - i)
  else if i <= 0
  then NO_NEXT_TRIGGER
  else next_array[i - 1];
next(i) = next_array[LENGTH - 1 - i]

@@

% Ignoring bounds checks, this roughly corresponds to @next(i) = if enabled(i) then i else next(i + 1)@.
% However, Kind2 does not support recursive functions; instead, we were able to express the recursion using Kind2's array comprehension syntax, which allows one to define an array as a function of its index.
% This encoding was further complicated by the fact that array comprehensions can only depend on values of the array with lower indices, not ones with a higher index: concretely, we cannot write @next[i] = if enabled(i) then i else next[i + 1]@; instead, we invert the indices: @next[i] = if enabled(LENGTH - 1 - i) then LENGTH - 1 - i else next[i - 1]@.

@@

\TODO{include comparison of required proof size vs runnable code}

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the simplified trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.

We were able to verify the Kind2 implementation of the simplified trigger-fetch mechanism for trigger arrays containing up to 32 elements; above that, our verification timed out after one hour.
For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.

>>>>

We were able to encode equivalent properties in Kind2 and in Pipit, aside from some encoding issues.
For example, the specification-only function that finds the next trigger is naturally recursive.
Kind2 does not support recursive functions, but we were able to encode it by introducing a temporary array and using Kind2's array comprehension syntax for scanning over arrays.
Additionally, while the recursive call \emph{increases} the index, the array scan can only depend on values with lower indices.
Below we compare a hypothetical unsupported recursive definition of @next@ (left) with the array-scan encoding (right):

\begin{minipage}{0.4\textwidth}
\begin{verbatim}
next(i) =
  if enabled(triggers,i)
  then i
  else if i >= LENGTH - 1
  then NO_NEXT_TRIGGER
  else next(i + 1);

\end{verbatim}
\end{minipage}
\begin{minipage}{0.55\textwidth}
\begin{verbatim}
next_array[i] =
  if enabled(triggers,LENGTH - 1 - i)
  then LENGTH - 1 - i
  else if i <= 0
  then NO_NEXT_TRIGGER
  else next_array[i - 1];
next = next_array[LENGTH - 1 - index];

@@

@@

\autoref{f:evaluation:kind2-runtime} shows the verification runtime for different sizes of arrays; the Pipit version is parametric in the array size, and is thus verified for all sizes of arrays.
We ran these experiments in Docker on an Intel i5-12500 with 32GB of RAM.
Both Kind2 and Pipit developments of the full trigger-fetch logic are roughly the same size, on the order of two-hundred lines of code including comments.
Ignoring whitespace and comments, the Pipit implementation has 26 lines of actual executable code, while the Kind2 code has 32.
The majority of the remaining code comprises the definition of valid schedules (34 for Pipit, 28 for Kind2), and the lemma statements and invariants (12 for Pipit, 31 for Kind2), as well as contract statements and boilerplate.

We were able to verify the Kind2 implementation of the complete trigger-fetch mechanism for up to 32 triggers; above that, our verification timed out after one hour.
For reference, the M_TTCAN hardware implementation of TTCAN supports up to 64 triggers~\cite{bosch2019mttcan}.
